---
title: "Aquaculture Exclusive Economic Zones"
date: November 26, 2025
author: "Sofia Rodas"
format: html
message: false
warning: false
---

```{r}
#|ouput: false

# load in necessary libraries
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
library(kableExtra)
```
## Read in data 
```{r}
#| output: false

# read in exclusive economic zones vector data
eez <- vect(here::here("data","wc_regions_clean.shp"))

# read in eez data as an sf object
eez_sf <- read_sf(here::here("data","wc_regions_clean.shp"))
```
```{r}
#| output: false
# read in raster depth data
depth <- rast(here::here("data", "depth.tif"))
```


```{r}
#| output: false

# read in raster sea surface temperature data
sst_2008 <- rast(here::here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here::here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here::here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here::here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here::here("data", "average_annual_sst_2012.tif"))

# stack data into a raster stack
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```
```{r}
# remove unnecessary raster data
rm(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

## Check CRS 
Use crs for rasters and st_crs for sf objects.
```{r}
#| warning: true

  if (crs(sst)== crs(depth)){
  print("it's a match!")
}else {
    warning("still not a match")
}

 if (crs(sst)== crs(eez)){
  print("it's a match!")
}else {
    warning("still not a match")
}

 if (crs(sst) == st_crs(eez_sf)){
  print("it's a match!")
}else {
    warning("still not a match")
}
```
Making all the CRS match the SST CRS.
```{r}
depth <- project(depth, crs(sst))
eez <- project(eez, crs(sst))

st_crs(eez_sf) <- crs(sst)
# eez_sf <- st_transform(eez_sf, crs(sst))
```

Make sure the CRS transformation was successful.
```{r}
#| warning: true

  if (crs(sst)== crs(depth)){
  print("it's a match!")
}else {
    warning("still not a match")
}

 if (crs(sst) == crs(eez)){
  print("it's a match!")
}else {
    warning("still not a match")
}

 if (crs(sst) == st_crs(eez_sf)){
  print("it's a match!")
}else {
    warning("still not a match")
}
```

## Data processing

This analysis works off the mean of SST measured between 2008 and 2012.
```{r}
# calculate the mean of sst 
mean_sst <- mean(sst)
# convert the units from kelvin to celcius
mean_sst <- mean_sst - 273 
```
This analysis requires SST and depth to have the same extent, resolution, and positions.
Crop to make sure the extent in the rasters are the same. 
```{r}
# crop the depth raster 
depth <- crop(depth, mean_sst)
```
Resample to make sure that the resolutions are the same.
```{r}
depth <- resample(depth, mean_sst)
```

Make sure the extents match.
```{r}
#| warning: true

  if (ext(mean_sst)== ext(depth)){
  print("extents match!")
}else {
    warning("extents not a match")
}

 
if (res(mean_sst)[1]== res(depth)[1]){
   print("resolution argument 1 is a match!")
 }else {
     warning("resolutions not a match")
 }

if (res(mean_sst)[2]== res(depth)[2]){
   print("resolution argument 2 is a match!")
 }else {
     warning("resolutions not a match")
 }
```

# Find cells where conditions are satisfied for both the sst and depth for specific species
Create true false for each 
1. Make a matrix
2. Reclassify based on that matrix

Sea surface temperature is adequate for oysters in the range of 11-30°C. Classify sea surface temperature for those that are adequate.
```{r}
# create a classification matrix for sst
sst_matrix <- matrix(c(-Inf, 11, 0, 
            11, 30, 1, 
            30, Inf, 0),
            ncol = 3, byrow = TRUE)

# classify where suitable sst habitat for oysters are
sst_classified <- classify(mean_sst, rcl = sst_matrix)
```

Mean meters below sea level (-70 - 0). 
```{r}
# create a classification matrix for depth
depth_matrix = matrix(c(-Inf, -70, 0,
                  -70, 0, 1,
                  0, Inf, 0),
                  ncol = 3, byrow = TRUE)

# classify where suitable depth habitat for oysters are
depth_classified <- classify(depth, rcl = depth_matrix)
```

Create a function to determine what locations are suited for oyster aquaculture based on both sea surface temperature and depth.
```{r}
suitable_habitat <- function(rast1, rast2){

  rast1 * rast2
}
```

```{r}
suit_oyster_aqu <- lapp(c(sst_classified, depth_classified), fun = suitable_habitat)
```

```{r}
# adequate depth for oyster aquaculture
tm_shape(sst_classified) +
  tm_raster() + 
  tm_title(text = "Adequate Sea Surface Temperatures for Oyster Aquaculture")

# adequate depth for oyster aquaculture
tm_shape(depth_classified) +
  tm_raster() + 
  tm_title(text = "Adequate Depths for Oyster Aquaculture")

# adequate locations for oyster aquaculture
tm_shape(suit_oyster_aqu) +
  tm_raster(col.scale = tm_scale_continuous(values = "Blues")) + 
  tm_title(text = "Adequate Locations for Oyster Aquaculture based on bathymetry and sea surface temperature") 
#tm_layout(legend.show = FALSE)
```

# We want to stay within Exclusive Economic Zones
We want to mask with the eez
```{r}
eez_rast <- rasterize(eez_sf, suit_oyster_aqu, field = "rgn_id")

suit_oyster_aqu[suit_oyster_aqu == 0] <- NA

eez_oysters <- mask(eez_rast, suit_oyster_aqu)
```
Then join the reclassified with the eez to get a dataframe with value, eez, x, y.
Count the cells. Number of cells that equal 1 times the cellsize() = total area that is suitable. 

```{r}
# zones = region
# summary = count()

oyst_count <- zonal(suit_oyster_aqu, eez_oysters, fun = "sum") |> 
  as.data.frame()

# oyster_aquaculture <- zonal(eez_oysters, eez_rast, fun = 'notNA') |> 
#   mutate(region = eez$rgn)

total_oyst <- expanse(eez_oysters)
```
#### How much suitable space is there for oyster aquaculture?
The total amount of space in the  Exclusive Economic Zones (EEZ) that are suitable for oyster aquaculture is `r total_oyst`.

```{r}
# number of cells that equal 1 * cellsize() = total area that is suitable
# expanse WOULD NEED TO MAKE THINGS NA 

avg_area <- cellSize(eez_oysters, unit = "km")

# avg_area <- mean(avg_area)

# area <- (oyster_aquaculture$lyr1 * avg_area) |> 
#   as.data.frame()

zonal_area <- zonal(avg_area, eez_oysters, fun = "sum") |>
  as.data.frame()
  
```

```{r}
eez_sf <-  left_join(eez_sf, oyst_count, by = "rgn_id")
eez_sf <-  left_join(eez_sf, zonal_area, by = "rgn_id")
```


```{r}
# tm_basemap("Esri.OceanBasemap") +
# tm_shape(eez_rast) +
#   tm_raster(col  = tm_vars("rgn"),
#               col.scale = tm_scale(values = (c("#7FB7BE",
#                                                 "#D3F3EE",
#                                                 "#5E6472",
#                                                 "#065A82",
#                                                 "#1F1A38"))),
#               fill.legend = tm_legend("West Coast EEZ Regions")) +
#   tm_shape(suit_oyster_aqu) + 
#   tm_raster(col.scale = tm_scale_categorical(labels = (c("Suitable Areas", "Unsuitable")),
#                                           values= (c("red", "white"))),
#             col.legend = tm_legend("Aquaculture Areas"))

```
```{r}
tm_basemap("Esri.OceanBasemap") +
tm_shape(eez_sf) +
  tm_polygons( fill = 'area',
              fill.legend = tm_legend("Suitable Aquaculture Area")) +
  tm_shape(suit_oyster_aqu) + 
  tm_raster(col.scale = tm_scale_categorical(labels = (c("Suitable Areas", "Unsuitable")),
                                          values= (c("red", "white"))),
            col.legend = tm_legend("Aquaculture Areas"))

```


```{r}
# regions <- c("Oregon", "Northern California", "Central California", "Southern California", "Washington")
# 
# # suit_oyster_area_df <- as.data.frame(suit_oyster_area)
# 
# eez_areas <- function(suit_oyster_aqu, eez_sf, regions){
#  for (regions in 1:5){
#    
#     eez_region_area <- subset(eez_sf, rgn == regions)
#   
#   # eez_region_area <- eez_sf[eez_sf$rgn == regions]
#   
#     suit_oyster_area <- mask(suit_oyster_aqu, eez_region_area)
#   
#     suit_oyster_area[suit_oyster_area == 0] <- NA
#   
#     suit_area <- expanse(suit_oyster_area)[1, 2]
#     
#     return(suit_area)
#     # suit_oyster_area_df <- as.data.frame(suit_oyster_area)
#     # 
#     #  suit_area <- suit_oyster_area_df |> 
#     #  mutate(area = expanse(suit_oyster_area)) 
#     # eez_sf <- eez_sf |> mutate(area = suit_area)
#  }
#   
# 
# }
```

```{r}
 # eez_region_area <- subset(eez_sf, rgn == "Washington")
 #  
 #  # eez_region_area <- eez_sf[eez_sf$rgn == regions]
 #  
 #    suit_oyster_area <- mask(suit_oyster_aqu, eez_region_area)
 #  
 #    suit_oyster_area[suit_oyster_area == 0] <- NA
 #  
 #    suit_area <- expanse(suit_oyster_area)[1, 2]
```


```{r}
# eez_areas(suit_oyster_aqu, eez_sf, regions)
```


```{r}
# create a table with the counts of adequate aquaculture zones 
oyster_aquaculture |> 
  relocate(region, .before = lyr1) |> 
  rename(counts = lyr1) |> 
  kbl(caption = "West Coast regions Exclusive Economic Zones (EEZ).") |> 
  kable_styling() 
```

# Writing a function
```{r}
#function (min_sst, max_sst, min_depth, max_depth)
```

# Additional species of interest: **Red Sea Urchin**, *Mesocentrotus franciscanus*
```{r}
# create a classification matrix for sst
urch_sst_matrix <- matrix(c(-Inf, 6.75, 0, 
            6.75, 13.02, 1, 
            13.02, Inf, 0),
            ncol = 3, byrow = TRUE)

# classify where suitable sst habitat for oysters are
urch_sst_classified <- classify(mean_sst, rcl = urch_sst_matrix)
```

```{r}
# create a classification matrix for depth
urch_depth_matrix = matrix(c(-Inf, -125, 0,
                  -125, 0, 1,
                  0, Inf, 0),
                  ncol = 3, byrow = TRUE)

# classify where suitable depth habitat for oysters are
urch_depth_classified <- classify(depth, rcl = urch_depth_matrix)
```

