---
title: "Aquaculture Exclusive Economic Zones"
date: November 26, 2025
author: "Sofia Rodas"
format: html
message: false
warning: false
---

```{r}
#|ouput: false

# load in necessary libraries
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
```
## Read in data 
```{r}
#| output: false

# read in exclusive economic zones vector data
eez <- vect(here::here("data","wc_regions_clean.shp"))
```
```{r}
#| output: false
# read in raster depth data
depth <- rast(here::here("data", "depth.tif"))
```


```{r}
#| output: false

# read in raster sea surface temperature data
sst_2008 <- rast(here::here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here::here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here::here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here::here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here::here("data", "average_annual_sst_2012.tif"))

# stack data into a raster stack
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```
```{r}
# remove unnecessary raster data
rm(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

## Check CRS 
Use crs for rasters and st_crs for sf objects.
```{r}
#| warning: true

  if (crs(sst)== crs(depth)){
  print("it's a match!")
}else {
    warning("still not a match")
}

 if (crs(sst)== crs(eez)){
  print("it's a match!")
}else {
    warning("still not a match")
}
```
Making all the CRS match the SST CRS.
```{r}
depth <- project(depth, crs(sst))
eez <- project(eez, crs(sst))
```

Make sure the CRS transformation was successful.
```{r}
#| warning: true

  if (crs(sst)== crs(depth)){
  print("it's a match!")
}else {
    warning("still not a match")
}

 if (crs(sst) == crs(eez)){
  print("it's a match!")
}else {
    warning("still not a match")
}
```

## Data processing

This analysis works off the mean of SST measured between 2008 and 2012.
```{r}
# calculate the mean of sst 
mean_sst <- mean(sst)
# convert the units from kelvin to celcius
mean_sst <- mean_sst - 273 
```
This analysis requires SST and depth to have the same extent, resolution, and positions.
Crop to make sure the extent in the rasters are the same. 
```{r}
# crop the depth raster 
depth <- crop(depth, mean_sst)
```
Resample to make sure that the resolutions are the same.
```{r}
depth <- resample(depth, mean_sst)
```

Make sure the extents match.
```{r}
#| warning: true

  if (ext(mean_sst)== ext(depth)){
  print("extents match!")
}else {
    warning("extents not a match")
}

 
if (res(mean_sst)[1]== res(depth)[1]){
   print("resolution argument 1 is a match!")
 }else {
     warning("resolutions not a match")
 }

if (res(mean_sst)[2]== res(depth)[2]){
   print("resolution argument 2 is a match!")
 }else {
     warning("resolutions not a match")
 }
```

# Find cells where conditions are satisfied for both the sst and depth for specific species
Create true false for each 
1. Make a matrix
2. Reclassify based on that matrix

Sea surface temperature is adequate for oysters in the range of 11-30°C. Classify sea surface temperature for those that are adequate.
```{r}
# create a classification matrix for sst
sst_matrix <- matrix(c(-Inf, 11, 0, 
            11, 30, 1, 
            30, Inf, 0),
            ncol = 3, byrow = TRUE)

# classify where suitable sst habitat for oysters are
sst_classified <- classify(mean_sst, rcl = sst_matrix)
```

Mean meters below sea level (-70 - 0). 
```{r}
# create a classification matrix for depth
depth_matrix = matrix(c(-Inf, -70, 0,
                  -70, 0, 1,
                  0, Inf, 0),
                  ncol = 3, byrow = TRUE)

# classify where suitable depth habitat for oysters are
depth_classified <- classify(depth, rcl = depth_matrix)
```

Create a function to determine what locations are suited for oyster aquaculture based on both sea surface temperature and depth.
```{r}
suitable_habitat <- function(rast1, rast2){
  
  rast1 * rast2
  
}
```

```{r}
suit_oyster_aqu <- lapp(c(sst_classified, depth_classified), fun = suitable_habitat)
```

```{r}
# adequate depth for oyster aquaculture
tm_shape(sst_classified) +
  tm_raster() + 
  tm_title(text = "Adequate Sea Surface Temperatures for Oyster Aquaculture")

# adequate depth for oyster aquaculture
tm_shape(depth_classified) +
  tm_raster() + 
  tm_title(text = "Adequate Depths for Oyster Aquaculture")

# adequate locations for oyster aquaculture
tm_shape(suit_oyster_aqu) +
  tm_raster(col.scale = tm_scale_continuous(values = "Blues")) + 
  tm_title(text = "Adequate Locations for Oyster Aquaculture based on bathymetry and sea surface temperature") 
#tm_layout(legend.show = FALSE)
```

# We want to stay within Exclusive Economic Zones
We want to mask with the eez
```{r}
eez_oysters <- mask(suit_oyster_aqu, eez)

eez_oysters[eez_oysters == 0] <- NA
```
Then join the reclassified with the eez to get a dataframe with value, eez, x, y.
Count the cells. Number of cells that equal 1 times the cellsize() = total area that is suitable. 

```{r}
# zones = region
# summary = count()

oyster_aquaculture <- zonal(eez_oysters, eez, fun = 'notNA')
```

```{r}
# number of cells that equal 1 * cellsize() = total area that is suitable
# expanse WOULD NEED TO MAKE THINGS NA 

cellSize(eez_oysters)
```

# Writing a function
```{r}
#function (min_sst, max_sst, min_depth, max_depth)
```

