---
title: "Aquaculture Exclusive Economic Zones"
date: November 26, 2025
author: "Sofia Rodas"
format: html
message: false
warning: false
---

```{r}
#|ouput: false

# load in necessary libraries
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
```
## Read in data 
```{r}
#| output: false

wc_eez <- st_read(here::here("data","wc_regions_clean.shp"))
```
```{r}
depth <- rast(here::here("data", "depth.tif"))
```


```{r}
# 
sst_2008 <- rast(here::here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here::here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here::here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here::here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here::here("data", "average_annual_sst_2012.tif"))

# stack data
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

## Function to Check CRS 
Use crs for rasters and st_crs for sf objects.
Keep the sst crs.
```{r}
match_crs <- function(df1, df2){
  if (crs(df1)==st_crs(df2)){
  print("it's a match!")
}else {
    warning("still not a match")
}}
```

```{r}
match_crs(sst, wc_eez)
```

```{r}
# for raster data to change crs
# project(st_crs(sst))
# for st data
# st_transform(st_crs)
```


```{r}
# to get the mean of the sst layers
# mean(sst)
# all of the cells are in kelvin (global function)
# sst - 273 

```

To stack the sst with the depth we need to make sure the CRS, extent, and resolution.
Crop to make sure the extent in the rasters are the same. 
```{r}
# need to have the same extent
# crop the depth raster 
# crop(depth, mean_sst)

```
Resample to make sure that the resolution is the same.
```{r}
# resample(depth, mean_sst)
```
Now we can join the mean and the depths
```{r}
# mean_sst , depth
# c(mean_sst, depth)
```

# Find cells where conditions are satisfied for both the sst and depth for specific species
Create true false for each 
1. Make a matrix
2. Reclassify based on that matrix

```{r}
# likely want to change false and true to 0 and 1 
# matrix = (-Inf , 11, FALSE, 
#            11, 30, TRUE, 
#            30, Inf, FALSE)

```

```{r}
# classify (mean_sst, matrix)
```

Mean meters below sea level (-70 - 0). Note: double check no oysters are above sea level.
```{r}
# zeros are false and 1 is true
# matrix = (-Inf, -70, 0
#           -70, 0, 1
#           0, Inf, 0)
```

We only want to keep where both of these are true. 
```{r}
# depth * sst
```

# We want to stay within Exclusive Economic Zones
We want to mask with the eez
```{r}
# mask(rcl, eez)
```
Then join the reclassified with the eez to get a dataframe with value, eez, x, y.
Count the cells. Number of cells that equal 1 times the cellsize() = total area that is suitable. 

```{r}
# zones = region
# summary = count()
```

```{r}
# number of cells that equal 1 * cellsize() = total area that is suitable
```

# Writing a function
```{r}
#function (min_sst, max_sst, min_depth, max_depth)
```

