---
title: "Potential Aquaculture Locations within Exclusive Economic Zones"
date: November 29, 2025
author: "Sofia Rodas"
format: 
  html: 
    code-fold: show
    embed-resources: true
execute: 
  echo: true
  message: false
  warning: false
  error: false
---

```{r}
#| label: import libraries
#| ouput: false

# load in necessary libraries
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
library(kableExtra)
```

## Read in data

```{r}
#| label: eez data read-in
#| output: false

# read in eez data as an sf object
eez_sf <- read_sf(here::here("data","wc_regions_clean.shp"))
```

```{r}
#| label: depth data read-in
#| output: false

# read in raster depth data
depth <- rast(here::here("data", "depth.tif"))
```

```{r}
#| label: sst data read-in
#| output: false

# read in raster sea surface temperature data
sst_2008 <- rast(here::here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here::here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here::here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here::here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here::here("data", "average_annual_sst_2012.tif"))

# stack data into a raster stack
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

```{r}
#| label: clean environment

# remove unnecessary raster data
rm(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

## Check CRS
Note: To check the CRS of a raster use crs. To check the CRS of a simple feature object use st_crs.
```{r}
#| label: crs check
#| warning: true

  if (crs(sst)== crs(depth)){
  print("it's a match!")
}else {
    warning("still not a match")
}
```

Make all of the CRS match the SST CRS.

```{r}
#| label: crs matching

depth <- project(depth, crs(sst))
eez_sf <- st_transform(eez_sf, crs(sst))
```

Make sure the CRS alterations were successful.

```{r}
#| label: crs match confirmation
#| warning: true

  if (crs(sst)== crs(depth)){
  print("it's a match!")
}else {
    warning("still not a match")
}

```

## Data processing

This analysis works off the mean of SST measured between 2008 and 2012.

```{r}
#| label: sst mean and celcius

# calculate the mean of sst 
mean_sst <- mean(sst)

# convert the units from kelvin to celcius
mean_sst <- mean_sst - 273 
```

This analysis requires SST and depth to have the same extent, resolution, and positions. Crop to make sure the extent of the rasters are the same.

```{r}
#| label: depth extent match

# crop the depth raster 
depth <- crop(depth, mean_sst)
```

Resample to make sure that the resolutions of the rasters are the same.

```{r}
#| label: depth resolution match

depth <- resample(depth, mean_sst)
```

Make sure the extents and resolutions match for the SST and depth data frames.

```{r}
#| label: depth extent and resolution confirmation
#| warning: true

  if (ext(mean_sst)== ext(depth)){
  print("extents match!")
}else {
    warning("extents not a match")
}

 
if (res(mean_sst)[1]== res(depth)[1] & res(mean_sst)[2]== res(depth)[2]){
   print("resolutions match!")
 }else {
     warning("resolutions not a match")
 }
```

# Find cells where conditions are satisfied for both the sst and depth for specific species.

1.  Make a matrix
2.  Reclassify based on that matrix

Sea surface temperature is adequate for oysters in the range of 11-30°C. Classify sea surface temperature to find those areas that are potential zones for oyster aquaculture.

```{r}
#| label: sst classification

# create a classification matrix for sst
sst_matrix <- matrix(c(-Inf, 11, 0, 
            11, 30, 1, 
            30, Inf, 0),
            ncol = 3, byrow = TRUE)

# classify where suitable sst habitat for oysters are
sst_classified <- classify(mean_sst, rcl = sst_matrix)
```

Oysters can survive between the depths of 70 and 0 below sea level. Classify depth to find those areas that are potential zones for oyster aquaculture.

```{r}
#| label: depth classification

# create a classification matrix for depth
depth_matrix = matrix(c(-Inf, -70, 0,
                  -70, 0, 1,
                  0, Inf, 0),
                  ncol = 3, byrow = TRUE)

# classify where suitable depth habitat for oysters are
depth_classified <- classify(depth, rcl = depth_matrix)
```

Create a function to determine what locations are suited for oyster aquaculture based on both sea surface temperature and depth.

```{r}
#| label: suitable area function 

suitable_habitat <- function(rast1, rast2){

  rast1 * rast2
}
```

Apply the function to determine what areas are suitable based on both sst and depth for oyster aquaculture.

```{r}
#| label: suitable area application

suit_oyster_aqu <- lapp(c(sst_classified, depth_classified), fun = suitable_habitat)

# set any locations with a zero to NA
suit_oyster_aqu[suit_oyster_aqu == 0] <- NA
```

Preliminary maps to assess differences between suitable locations based on depth, sst, and both depth and sst.

```{r}
#| label: initial exploration maps

# suitable depth for oyster aquaculture
tm_basemap("Esri.OceanBasemap") +
tm_shape(sst_classified) +
  tm_raster(style = "cat",
            palette = c("skyblue", "red"),
            labels = c("unsuitable", "suitable"),
            title = "Sea Surface Temperature (SST)") + 
  tm_title(text = "Suitable Sea Surface Temperatures for Oyster Aquaculture",
           size = 2) +
  tm_compass() +
  tm_scalebar() +
  tm_layout(fontsize = 2,
            scale = 0.5,
            compass.position = tm_pos_in("left", "bottom"),
            scalebar.position = tm_pos_in("left", "bottom")
            ) 

# suitable depth for oyster aquaculture
tm_shape(depth_classified) +
  tm_raster(style = "cat",
            palette = c("skyblue", "red"),
            labels = c("unsuitable", "suitable"),
            title = "Depth suitability") + 
  tm_title(text = "Suitable Depths for Oyster Aquaculture") + 
  tm_compass() +
  tm_scalebar() +
  tm_layout(
            compass.position = tm_pos_in("right", "bottom"),
            scalebar.position = tm_pos_out("center", "bottom")
            ) 

# suitable locations for oyster aquaculture
tm_basemap("Esri.OceanBasemap") +
tm_shape(suit_oyster_aqu) +
  tm_raster(style = "cat",
            palette = c( "red"),
            labels = c("suitable"),
            title = "Aquaculture suitability") + 
  tm_title(text = "Suitable Locations for Oyster Aquaculture based on bathymetry and sea surface temperature") +
  tm_compass() +
  tm_scalebar() +
  tm_layout(
            compass.position = tm_pos_out("right", "bottom"),
            scalebar.position = tm_pos_out("center", "bottom")
            ) 


```

# How much suitable aquaculture space is there in each Exclusive Economic Zone (EEZ)?

Find the suitable aquaculture locations for each EEZ.

```{r}
#| label: suitable area per eez

# rasterize the eez sf object
eez_rast <- rasterize(eez_sf, suit_oyster_aqu, field = "rgn_id")

# set any locations with a zero to NA
suit_oyster_aqu[suit_oyster_aqu == 0] <- NA

# select only suitable oyster aquaculture locations that are within the eez
eez_oysters <- mask(suit_oyster_aqu, eez_sf)

# calculate average size of cells
avg_area <- cellSize(eez_oysters, unit = "km")

# sum the suitable aquaculture locations based on eez zones
oyst_area <- zonal(x = (eez_oysters * avg_area), 
                   z = eez_rast, 
                   fun = "sum", 
                   na.rm = TRUE) |> 
  as.data.frame()

# join to make one data frame with the total area
eez_sf <-  left_join(eez_sf, oyst_area, by = "rgn_id")

# rename new column 
eez_sf <- eez_sf |> 
  rename("area" = "lyr1")
```

#### How much suitable space is there for oyster aquaculture?
```{r}
total_oyst <- sum(eez_sf$area)
```

The total amount of space in the Exclusive Economic Zones (EEZ) on the West Coast that are suitable for oyster aquaculture based on suitable sea surface temperature and suitable bathymetry is `r total_oyst`.

```{r}
#| label: eez map by suitable area amount

tm_basemap("Esri.OceanBasemap") +
tm_shape(eez_sf) +
  tm_polygons( fill = 'area',
              fill.legend = tm_legend("Suitable Aquaculture Area")) +
  tm_text("rgn", size = 0.7) +
  tm_shape(eez_oysters) + 
  tm_raster(col.scale = tm_scale_categorical(labels = (c("Suitable Areas", "Unsuitable")),
                                          values= (c("red", "white"))),
            col.legend = tm_legend("Aquaculture Areas")) + 
  tm_compass() +
  tm_scalebar() +
  tm_title("Exclusive Economic Zones shaded by Suitable Oyster Aquaculture Area ") +
  tm_layout(compass.position = tm_pos_out("right", "bottom"),
            scalebar.position = tm_pos_out("center", "bottom")
            ) 
# CHANGE LEGEND SIZE
# ADD HARD BREAKS FOR OYSTERS 

```


```{r}
# create a table with the counts of adequate aquaculture zones
# oyster_aquaculture |>
#   relocate(region, .before = lyr1) |>
#   rename(counts = lyr1) |>
#   kbl(caption = "West Coast regions Exclusive Economic Zones (EEZ).") |>
#   kable_styling()
```

# Potential aquaculture locations

```{r}
#| label: potential aquaculture location function

potent_aqu <- function (sst, 
                        eez_sf, 
                        depth, 
                        min_sst, 
                        max_sst, 
                        min_depth, 
                        max_depth, 
                        species_name){

# make all of the CRS match
depth <- project(depth, crs(sst))
st_crs(eez_sf) <- crs(sst)

# calculate the mean of sst 
mean_sst <- mean(sst)
# convert the units from kelvin to celcius
mean_sst <- mean_sst - 273 

# crop the depth raster to match the sst extent
depth <- crop(depth, mean_sst)

# resample depth raster to match the sst resolution
depth <- resample(depth, mean_sst)

# create a classification matrix for sst
sst_matrix <- matrix(c(-Inf, min_sst, 0, 
            min_sst, max_sst, 1, 
            max_sst, Inf, 0),
            ncol = 3, byrow = TRUE)

# classify where suitable sst habitat for oysters are
sst_classified <- classify(mean_sst, rcl = sst_matrix)

# create a classification matrix for depth
depth_matrix = matrix(c(-Inf, min_depth, 0,
                  min_depth, max_depth, 1,
                  max_depth, Inf, 0),
                  ncol = 3, byrow = TRUE)

# classify where suitable depth habitat for oysters are
depth_classified <- classify(depth, rcl = depth_matrix)

# function to determine what habitats are suitable 
suitable_habitat <- function(rast1, rast2){

  rast1 * rast2
}

# find suitable locations for aquaculture
suit_aqu <- lapp(c(sst_classified, depth_classified), fun = suitable_habitat)

# rasterize the eez sf object
eez_rast <- rasterize(eez_sf, suit_aqu, field = "rgn_id")

# set any locations with a zero to NA
suit_aqu[suit_aqu == 0] <- NA

# select only suitable oyster aquaculture locations that are within the eez
eez_spec <- mask(suit_aqu, eez_sf)

# calculate average size of cells
avg_area <- cellSize(eez_spec, unit = "km")

# sum the suitable aquaculture locations based on eez zones
spec_area <- zonal(x = (eez_spec * avg_area), 
                   z = eez_rast, 
                   fun = "sum", 
                   na.rm = TRUE) |> 
              as.data.frame()

# join to make one data frame with the total area
spec_area_eez <-  left_join(eez_sf, spec_area, by = "rgn_id")

# rename new column 
spec_area_eez <- spec_area_eez |> 
  rename("spec_area" = "lyr1")

# map the eez's based on the amount of suitable area per region
tm_basemap("Esri.OceanBasemap") +
tm_shape(spec_area_eez) +
  tm_polygons(fill = "spec_area",
              fill.legend = tm_legend("Suitable Aquaculture Area")) +
  tm_text("rgn", size = 0.7) +
  tm_shape(eez_oysters) + 
  tm_raster(col.scale = tm_scale_categorical(labels = (c("Suitable Areas", "Unsuitable")),
                                          values= (c("red", "white"))),
            col.legend = tm_legend("Aquaculture Areas")) + 
  tm_compass() +
  tm_scalebar() +
  tm_title("Exclusive Economic Zones shaded by Suitable Oyster Aquaculture Area ") +
  tm_layout(compass.position = tm_pos_out("right", "bottom"),
            scalebar.position = tm_pos_out("center", "bottom")
            ) 
}
```
# Additional species of interest: **Red Sea Urchin**, *Mesocentrotus franciscanus*
```{r}
#| label: suitable area for Mesocentrotus franciscanus
potent_aqu(sst, eez_sf, depth, 6.75, 13.02, -125, 0, "Mesocentrotus franciscanus")
```

# References 
```{r}

```

